---
title: "Extract Acute HF Cohort (MIMIC-IV, DuckDB)"
author: "Zihan Xiong"
date: "2026-04-20"
output: html_document
---

```{r setup, message=FALSE}
library(duckdb)
library(DBI)
library(dplyr)

data_dir <- "data"
con <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")
```

# Register MIMIC tables as DuckDB views

```{r register-views}
register_view <- function(con, name, file) {
  path <- file.path(data_dir, file)
  sql <- paste0(
    "CREATE OR REPLACE VIEW ", name,
    " AS SELECT * FROM read_csv_auto('", path, "')"
  )
  dbExecute(con, sql)
}

register_view(con, "diagnoses_icd", "diagnoses_icd.csv.gz")
register_view(con, "admissions",    "admissions.csv.gz")
register_view(con, "patients",      "patients.csv.gz")
register_view(con, "icustays",      "icustays.csv.gz")
register_view(con, "d_labitems",    "d_labitems.csv.gz")
register_view(con, "labevents",     "labevents.csv.gz")
register_view(con, "d_items",       "d_items.csv.gz")
register_view(con, "chartevents",   "chartevents.csv.gz")
```

# Pull all acute HF patients (ICD codes from Yan et al. 2023)

```{r extract-hf}
hf_sql <- "
WITH acute_hf_codes AS (
  SELECT icd_code, icd_version, hadm_id, subject_id
  FROM diagnoses_icd
  WHERE (icd_version = 9 AND icd_code IN (
      '42821','42822','42823',
      '42831','42832','42833',
      '42841','42842','42843'
  ))
  OR (icd_version = 10 AND icd_code IN (
      'I5021','I5022','I5023',
      'I5031','I5032','I5033',
      'I5041','I5042','I5043',
      'I50811','I50812','I50813'
  ))
),
hf_subtype AS (
  SELECT
    hadm_id,
    subject_id,
    MAX(CASE
      WHEN icd_code IN ('42821','42822','42823','I5021','I5022','I5023')
      THEN 1 ELSE 0 END) AS is_hfref,
    MAX(CASE
      WHEN icd_code IN ('42831','42832','42833','I5031','I5032','I5033')
      THEN 1 ELSE 0 END) AS is_hfpef,
    MAX(CASE
      WHEN icd_code IN ('42841','42842','42843',
                        'I5041','I5042','I5043',
                        'I50811','I50812','I50813')
      THEN 1 ELSE 0 END) AS is_hfmixed
  FROM acute_hf_codes
  GROUP BY hadm_id, subject_id
)
SELECT
  h.subject_id,
  h.hadm_id,
  a.admittime,
  a.dischtime,
  a.deathtime,
  a.hospital_expire_flag,
  a.admission_type,
  a.race,
  a.insurance,
  p.gender,
  p.anchor_age,
  p.anchor_year,
  h.is_hfref,
  h.is_hfpef,
  h.is_hfmixed
FROM hf_subtype h
LEFT JOIN admissions a USING (hadm_id, subject_id)
LEFT JOIN patients p   USING (subject_id)
ORDER BY h.subject_id, a.admittime;
"

hf_cohort <- dbGetQuery(con, hf_sql)
```

# Summary counts

```{r summary}
cat("Acute HF admissions:", nrow(hf_cohort), "\n")
cat("Unique patients:    ", dplyr::n_distinct(hf_cohort$subject_id), "\n\n")

subtype_counts <- hf_cohort |>
  summarise(
    hfref  = sum(is_hfref,  na.rm = TRUE),
    hfpef  = sum(is_hfpef,  na.rm = TRUE),
    mixed  = sum(is_hfmixed, na.rm = TRUE)
  )
print(subtype_counts)

head(hf_cohort)
```

# Check AF patients within the HF cohort

```{r af-check}
dbWriteTable(con, "hf_cohort_ids",
             hf_cohort["hadm_id"], overwrite = TRUE)

af_sql <- "
WITH af_dx AS (
  SELECT DISTINCT subject_id, hadm_id
  FROM diagnoses_icd
  WHERE (icd_version = 9  AND icd_code = '42731')
     OR (icd_version = 10 AND icd_code IN ('I480','I481','I482','I4891'))
)
SELECT
  COUNT(DISTINCT h.hadm_id) AS n_hf_admissions,
  COUNT(DISTINCT CASE WHEN af.hadm_id IS NOT NULL
                      THEN h.hadm_id END) AS n_hf_with_af,
  ROUND(100.0 * COUNT(DISTINCT CASE WHEN af.hadm_id IS NOT NULL
                                    THEN h.hadm_id END)
              / COUNT(DISTINCT h.hadm_id), 2) AS pct_af
FROM hf_cohort_ids h
LEFT JOIN af_dx af USING (hadm_id);
"

af_counts <- dbGetQuery(con, af_sql)
print(af_counts)
```

# Restrict to ICU stays and re-check AF

```{r icu-af-check}
icu_af_sql <- "
WITH hf_icu AS (
  SELECT DISTINCT icu.subject_id, icu.hadm_id
  FROM icustays icu
  INNER JOIN hf_cohort_ids h USING (hadm_id)
),
af_dx AS (
  SELECT DISTINCT subject_id, hadm_id
  FROM diagnoses_icd
  WHERE (icd_version = 9  AND icd_code = '42731')
     OR (icd_version = 10 AND icd_code IN ('I480','I481','I482','I4891'))
)
SELECT
  COUNT(DISTINCT hf_icu.hadm_id) AS n_icu_hf,
  COUNT(DISTINCT hf_icu.subject_id) AS n_icu_hf_patients,
  COUNT(DISTINCT CASE WHEN af.hadm_id IS NOT NULL
                      THEN hf_icu.hadm_id END) AS n_icu_hf_with_af,
  ROUND(100.0 * COUNT(DISTINCT CASE WHEN af.hadm_id IS NOT NULL
                                    THEN hf_icu.hadm_id END)
              / COUNT(DISTINCT hf_icu.hadm_id), 2) AS pct_af
FROM hf_icu
LEFT JOIN af_dx af USING (hadm_id);
"

icu_af_counts <- dbGetQuery(con, icu_af_sql)
print(icu_af_counts)
```

# Build final cohort with exclusions

Apply (in order): demographics join, age >= 18, ICU >= 24h,
exclude AMA/hospice, first qualifying admission per patient,
exclude patients with prior AF.

```{r build-cohort}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_acute_hf AS
SELECT DISTINCT hadm_id, subject_id
FROM diagnoses_icd
WHERE (icd_version = 9  AND icd_code IN (
    '42821','42822','42823',
    '42831','42832','42833',
    '42841','42842','42843'))
   OR (icd_version = 10 AND icd_code IN (
    'I5021','I5022','I5023',
    'I5031','I5032','I5033',
    'I5041','I5042','I5043',
    'I50811','I50812','I50813'));
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_hf_admits AS
SELECT
  a.subject_id, a.hadm_id,
  a.admittime, a.dischtime, a.deathtime,
  a.hospital_expire_flag, a.discharge_location,
  a.race, a.insurance,
  p.gender, p.anchor_age, p.anchor_year,
  (p.anchor_age +
    EXTRACT(YEAR FROM a.admittime) - p.anchor_year) AS age_at_admit
FROM stg_acute_hf h
INNER JOIN admissions a USING (hadm_id, subject_id)
INNER JOIN patients   p USING (subject_id);
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_icu_first AS
SELECT
  subject_id, hadm_id,
  MIN(intime)  AS icu_intime,
  MAX(outtime) AS icu_outtime,
  SUM(los)     AS icu_total_los
FROM icustays
GROUP BY subject_id, hadm_id;
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_qualifying AS
SELECT h.*,
  icu.icu_intime, icu.icu_outtime, icu.icu_total_los
FROM stg_hf_admits h
INNER JOIN stg_icu_first icu USING (subject_id, hadm_id)
WHERE h.age_at_admit >= 18
  AND icu.icu_total_los >= 1.0
  AND h.dischtime IS NOT NULL
  AND (h.discharge_location IS NULL
       OR h.discharge_location NOT IN
           ('AGAINST ADVICE','HOSPICE'));
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_index AS
SELECT * FROM (
  SELECT q.*,
    ROW_NUMBER() OVER (
      PARTITION BY subject_id ORDER BY admittime
    ) AS rn
  FROM stg_qualifying q
) WHERE rn = 1;
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_af_dx AS
SELECT DISTINCT subject_id, hadm_id
FROM diagnoses_icd
WHERE (icd_version = 9  AND icd_code = '42731')
   OR (icd_version = 10 AND icd_code IN
           ('I480','I481','I482','I4891'));
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_prior_af AS
SELECT DISTINCT i.subject_id
FROM stg_index i
INNER JOIN admissions a2
  ON a2.subject_id = i.subject_id
INNER JOIN stg_af_dx af
  ON af.hadm_id = a2.hadm_id
WHERE a2.admittime < i.admittime;
")

dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_final AS
SELECT i.*,
  CASE WHEN af.hadm_id IS NOT NULL
       THEN 1 ELSE 0 END AS has_af_index
FROM stg_index i
LEFT JOIN stg_af_dx af ON af.hadm_id = i.hadm_id
WHERE i.subject_id NOT IN (
  SELECT subject_id FROM stg_prior_af
);
")

final_cohort <- dbGetQuery(con, "SELECT * FROM stg_final;")
cat("Final cohort rows:", nrow(final_cohort), "\n")
```

# Cohort flow diagram

```{r cohort-flow}
flow_counts <- dbGetQuery(con, "
SELECT 1 AS step_no,
       'acute HF admissions'            AS step,
       COUNT(DISTINCT hadm_id)          AS n_admits,
       COUNT(DISTINCT subject_id)       AS n_patients
FROM stg_acute_hf
UNION ALL
SELECT 2, 'joined demographics',
       COUNT(DISTINCT hadm_id),
       COUNT(DISTINCT subject_id)
FROM stg_hf_admits
UNION ALL
SELECT 3, 'age>=18, ICU>=24h, not AMA/hospice, has dischtime',
       COUNT(DISTINCT hadm_id),
       COUNT(DISTINCT subject_id)
FROM stg_qualifying
UNION ALL
SELECT 4, 'first qualifying admission per patient',
       COUNT(DISTINCT hadm_id),
       COUNT(DISTINCT subject_id)
FROM stg_index
UNION ALL
SELECT 5, 'final (no prior AF)',
       COUNT(DISTINCT hadm_id),
       COUNT(DISTINCT subject_id)
FROM stg_final
ORDER BY step_no;
")
print(flow_counts)
```

# AF event count + decision rule (PROJECT_CONTEXT §4.3)

```{r af-events}
af_event <- dbGetQuery(con, "
SELECT
  SUM(has_af_index)             AS n_af_events,
  COUNT(*)                      AS n_total,
  ROUND(100.0 * SUM(has_af_index)
        / COUNT(*), 2)          AS pct_af,
  SUM(hospital_expire_flag)     AS n_deaths
FROM stg_final;
")
print(af_event)

n_events <- af_event$n_af_events
if (n_events < 500) {
  cat("Decision: <500 AF events -> drop XGBoost, use LASSO\n")
} else if (n_events < 1000) {
  cat("Decision: 500-1000 -> XGBoost OK overall; ",
      "subgroup SHAP descriptive only\n", sep = "")
} else {
  cat("Decision: >=1000 -> full plan, XGBoost + subgroup SHAP\n")
}
```

## Secondary outcome: 30-day all-cause mortality

First, check `dod` coverage in the final cohort to decide whether
30-day post-discharge mortality is feasible as a secondary outcome.

```{r dod-coverage}
dod_check <- dbGetQuery(con, "
SELECT
  COUNT(*)                                       AS n_total,
  SUM(CASE WHEN p.dod IS NOT NULL THEN 1 ELSE 0 END)
                                                 AS n_with_dod,
  SUM(CASE WHEN f.hospital_expire_flag = 1
           THEN 1 ELSE 0 END)                    AS n_inhosp_death,
  SUM(CASE WHEN p.dod IS NOT NULL
           AND f.hospital_expire_flag = 0
           THEN 1 ELSE 0 END)                    AS n_post_discharge_death,
  SUM(CASE WHEN p.dod IS NOT NULL
           AND f.hospital_expire_flag = 0
           AND p.dod <= f.dischtime + INTERVAL 30 DAY
           THEN 1 ELSE 0 END)                    AS n_30d_postdc_death
FROM stg_final f
LEFT JOIN patients p USING (subject_id);
")
print(dod_check)
```

If `n_post_discharge_death` is large enough (ideally ≥ 200), build the
30-day mortality outcome and time-to-event variables for the secondary
Cox analysis.

```{r build-30d-outcome}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_final_surv AS
SELECT
  f.*,
  p.dod,
  CASE
    WHEN f.hospital_expire_flag = 1 THEN 1
    WHEN p.dod IS NOT NULL
         AND p.dod <= f.admittime + INTERVAL 30 DAY THEN 1
    ELSE 0
  END AS death_30d,
  CASE
    WHEN f.hospital_expire_flag = 1
      THEN DATE_DIFF('day', f.admittime, f.deathtime)
    WHEN p.dod IS NOT NULL
         AND p.dod <= f.admittime + INTERVAL 30 DAY
      THEN DATE_DIFF('day', f.admittime, p.dod)
    ELSE 30
  END AS time_to_event_30d,
  CASE
    WHEN f.hospital_expire_flag = 1 THEN 0
    WHEN p.dod IS NOT NULL
         AND p.dod <= f.admittime + INTERVAL 30 DAY THEN 0
    ELSE 1
  END AS censored_30d
FROM stg_final f
LEFT JOIN patients p USING (subject_id);
")

surv_summary <- dbGetQuery(con, "
SELECT
  COUNT(*)                          AS n,
  SUM(death_30d)                    AS n_deaths_30d,
  SUM(censored_30d)                 AS n_censored,
  ROUND(100.0 * SUM(death_30d) / COUNT(*), 2)
                                    AS pct_death_30d,
  ROUND(AVG(time_to_event_30d), 1)  AS mean_follow_days
FROM stg_final_surv;
")
print(surv_summary)
```

# Baseline covariates: HF subtype, comorbidities, first ICU unit

## HF subtype flag (recover from index admission diagnoses)

```{r extract-subtype}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_subtype AS
SELECT
  d.subject_id, d.hadm_id,
  MAX(CASE
    WHEN (d.icd_version = 9
          AND d.icd_code IN ('42821','42822','42823'))
      OR (d.icd_version = 10
          AND d.icd_code IN ('I5021','I5022','I5023'))
    THEN 1 ELSE 0 END)                                 AS is_hfref,
  MAX(CASE
    WHEN (d.icd_version = 9
          AND d.icd_code IN ('42831','42832','42833'))
      OR (d.icd_version = 10
          AND d.icd_code IN ('I5031','I5032','I5033'))
    THEN 1 ELSE 0 END)                                 AS is_hfpef,
  MAX(CASE
    WHEN (d.icd_version = 9
          AND d.icd_code IN ('42841','42842','42843'))
      OR (d.icd_version = 10
          AND d.icd_code IN ('I5041','I5042','I5043',
                             'I50811','I50812','I50813'))
    THEN 1 ELSE 0 END)                                 AS is_hfmixed
FROM diagnoses_icd d
INNER JOIN stg_final f USING (subject_id, hadm_id)
GROUP BY d.subject_id, d.hadm_id;
")
```

## Comorbidity binary flags (from all admissions up to and incl. index)

```{r extract-comorb}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_comorb AS
WITH pt_dx AS (
  SELECT
    f.subject_id,
    f.hadm_id AS index_hadm_id,
    d.icd_code,
    d.icd_version
  FROM stg_final f
  INNER JOIN admissions a
    ON a.subject_id = f.subject_id
   AND a.admittime <= f.admittime
  INNER JOIN diagnoses_icd d
    ON d.hadm_id     = a.hadm_id
   AND d.subject_id  = a.subject_id
)
SELECT
  subject_id,
  index_hadm_id AS hadm_id,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '401%'
                        OR icd_code LIKE '402%'
                        OR icd_code LIKE '403%'
                        OR icd_code LIKE '404%'
                        OR icd_code LIKE '405%'))
    OR (icd_version = 10 AND (icd_code LIKE 'I10%'
                           OR icd_code LIKE 'I11%'
                           OR icd_code LIKE 'I12%'
                           OR icd_code LIKE 'I13%'
                           OR icd_code LIKE 'I15%'))
    THEN 1 ELSE 0 END)                                AS has_htn,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '249%'
                        OR icd_code LIKE '250%'))
    OR (icd_version = 10 AND (icd_code LIKE 'E08%'
                           OR icd_code LIKE 'E09%'
                           OR icd_code LIKE 'E10%'
                           OR icd_code LIKE 'E11%'
                           OR icd_code LIKE 'E13%'))
    THEN 1 ELSE 0 END)                                AS has_dm,
  MAX(CASE WHEN
    (icd_version = 9  AND icd_code LIKE '585%')
    OR (icd_version = 10 AND icd_code LIKE 'N18%')
    THEN 1 ELSE 0 END)                                AS has_ckd,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '410%'
                        OR icd_code LIKE '411%'
                        OR icd_code LIKE '412%'
                        OR icd_code LIKE '413%'
                        OR icd_code LIKE '414%'))
    OR (icd_version = 10 AND (icd_code LIKE 'I20%'
                           OR icd_code LIKE 'I21%'
                           OR icd_code LIKE 'I22%'
                           OR icd_code LIKE 'I23%'
                           OR icd_code LIKE 'I24%'
                           OR icd_code LIKE 'I25%'))
    THEN 1 ELSE 0 END)                                AS has_cad,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '491%'
                        OR icd_code LIKE '492%'
                        OR icd_code LIKE '496%'))
    OR (icd_version = 10 AND (icd_code LIKE 'J41%'
                           OR icd_code LIKE 'J42%'
                           OR icd_code LIKE 'J43%'
                           OR icd_code LIKE 'J44%'))
    THEN 1 ELSE 0 END)                                AS has_copd,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '394%'
                        OR icd_code LIKE '395%'
                        OR icd_code LIKE '396%'
                        OR icd_code LIKE '397%'
                        OR icd_code LIKE '424%'))
    OR (icd_version = 10 AND (icd_code LIKE 'I05%'
                           OR icd_code LIKE 'I06%'
                           OR icd_code LIKE 'I07%'
                           OR icd_code LIKE 'I08%'
                           OR icd_code LIKE 'I09%'
                           OR icd_code LIKE 'I34%'
                           OR icd_code LIKE 'I35%'
                           OR icd_code LIKE 'I36%'
                           OR icd_code LIKE 'I37%'
                           OR icd_code LIKE 'I38%'
                           OR icd_code LIKE 'I39%'))
    THEN 1 ELSE 0 END)                                AS has_valvular,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '440%'
                        OR icd_code LIKE '4439%'))
    OR (icd_version = 10 AND (icd_code LIKE 'I70%'
                           OR icd_code LIKE 'I73%'))
    THEN 1 ELSE 0 END)                                AS has_pad,
  MAX(CASE WHEN
    (icd_version = 9  AND (icd_code LIKE '433%'
                        OR icd_code LIKE '434%'
                        OR icd_code LIKE '435%'
                        OR icd_code LIKE '436%'
                        OR icd_code LIKE '437%'
                        OR icd_code LIKE '438%'))
    OR (icd_version = 10 AND (icd_code LIKE 'I60%'
                           OR icd_code LIKE 'I61%'
                           OR icd_code LIKE 'I62%'
                           OR icd_code LIKE 'I63%'
                           OR icd_code LIKE 'I64%'
                           OR icd_code LIKE 'I65%'
                           OR icd_code LIKE 'I66%'
                           OR icd_code LIKE 'I67%'
                           OR icd_code LIKE 'I68%'
                           OR icd_code LIKE 'I69%'
                           OR icd_code LIKE 'G45%'))
    THEN 1 ELSE 0 END)                                AS has_stroke_hx,
  MAX(CASE WHEN
    (icd_version = 9  AND icd_code IN ('4160','4168','4169'))
    OR (icd_version = 10 AND icd_code LIKE 'I27%')
    THEN 1 ELSE 0 END)                                AS has_pulm_htn,
  MAX(CASE WHEN
    (icd_version = 9  AND icd_code LIKE '425%')
    OR (icd_version = 10 AND icd_code LIKE 'I42%')
    THEN 1 ELSE 0 END)                                AS has_cardiomyopathy
FROM pt_dx
GROUP BY subject_id, index_hadm_id;
")
```

## First care unit of the index ICU admission

```{r extract-careunit}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_careunit AS
SELECT subject_id, hadm_id, first_careunit
FROM icustays
QUALIFY ROW_NUMBER() OVER (
  PARTITION BY subject_id, hadm_id ORDER BY intime
) = 1;
")
```

## Merge into analytic dataset

```{r build-analytic}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_analytic AS
SELECT
  f.*,
  s.is_hfref, s.is_hfpef, s.is_hfmixed,
  c.has_htn, c.has_dm, c.has_ckd, c.has_cad, c.has_copd,
  c.has_valvular, c.has_pad, c.has_stroke_hx,
  c.has_pulm_htn, c.has_cardiomyopathy,
  u.first_careunit,
  CASE
    WHEN u.first_careunit IN (
           'Coronary Care Unit (CCU)',
           'Cardiac Vascular Intensive Care Unit (CVICU)')
      THEN 'Cardiac'
    WHEN u.first_careunit IN (
           'Medical Intensive Care Unit (MICU)',
           'Medical/Surgical Intensive Care Unit (MICU/SICU)')
      THEN 'Medical'
    ELSE 'Surgical/Other'
  END                                        AS careunit_cat
FROM stg_final f
LEFT JOIN stg_subtype  s USING (subject_id, hadm_id)
LEFT JOIN stg_comorb   c USING (subject_id, hadm_id)
LEFT JOIN stg_careunit u USING (subject_id, hadm_id);
")

analytic <- dbGetQuery(con, "SELECT * FROM stg_analytic;")
cat("Rows:", nrow(analytic),
    "  Columns:", ncol(analytic), "\n")
str(analytic)
```

## Quick Table 1 snapshot

```{r table1-snapshot}
summary_tab <- dbGetQuery(con, "
SELECT
  COUNT(*)                                      AS n,
  ROUND(AVG(age_at_admit), 1)                   AS mean_age,
  ROUND(100.0 * SUM(CASE WHEN gender = 'F' THEN 1 ELSE 0 END)
        / COUNT(*), 1)                          AS pct_female,
  ROUND(100.0 * SUM(is_hfref) / COUNT(*), 1)    AS pct_hfref,
  ROUND(100.0 * SUM(is_hfpef) / COUNT(*), 1)    AS pct_hfpef,
  ROUND(100.0 * SUM(is_hfmixed) / COUNT(*), 1)  AS pct_hfmixed,
  ROUND(100.0 * SUM(has_htn) / COUNT(*), 1)            AS pct_htn,
  ROUND(100.0 * SUM(has_dm) / COUNT(*), 1)             AS pct_dm,
  ROUND(100.0 * SUM(has_ckd) / COUNT(*), 1)            AS pct_ckd,
  ROUND(100.0 * SUM(has_cad) / COUNT(*), 1)            AS pct_cad,
  ROUND(100.0 * SUM(has_copd) / COUNT(*), 1)           AS pct_copd,
  ROUND(100.0 * SUM(has_valvular) / COUNT(*), 1)       AS pct_valvular,
  ROUND(100.0 * SUM(has_pad) / COUNT(*), 1)            AS pct_pad,
  ROUND(100.0 * SUM(has_stroke_hx) / COUNT(*), 1)      AS pct_stroke,
  ROUND(100.0 * SUM(has_pulm_htn) / COUNT(*), 1)       AS pct_pulmhtn,
  ROUND(100.0 * SUM(has_cardiomyopathy) / COUNT(*), 1) AS pct_cmp
FROM stg_analytic;
")
print(summary_tab)

careunit_dist <- dbGetQuery(con, "
SELECT first_careunit, COUNT(*) AS n,
       ROUND(100.0 * COUNT(*) /
             (SELECT COUNT(*) FROM stg_analytic), 1) AS pct
FROM stg_analytic
GROUP BY first_careunit
ORDER BY n DESC;
")
print(careunit_dist)

careunit_cat_dist <- dbGetQuery(con, "
SELECT careunit_cat, COUNT(*) AS n,
       ROUND(100.0 * COUNT(*) /
             (SELECT COUNT(*) FROM stg_analytic), 1) AS pct
FROM stg_analytic
GROUP BY careunit_cat
ORDER BY n DESC;
")
print(careunit_cat_dist)
```

# First-24h physiology: labs

Extract extremes (MIN / MAX) of 10 labs within 24h of ICU intime.
Source: `labevents` + `d_labitems` lookup. Keyed on (subject_id, hadm_id).

## Lab itemid lookup (for reference)

```{r explore-labitems}
labitem_lookup <- dbGetQuery(con, "
SELECT itemid, label, fluid, category
FROM d_labitems
WHERE fluid = 'Blood'
  AND label IN ('White Blood Cells',
                'Hemoglobin',
                'Platelet Count',
                'Creatinine',
                'Urea Nitrogen',
                'Sodium',
                'Potassium',
                'Glucose',
                'Anion Gap',
                'RDW')
ORDER BY label;
")
print(labitem_lookup)
```

## Extract 24h lab extremes

```{r extract-labs-24h}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_labs_24h AS
WITH lab_items AS (
  SELECT itemid, label
  FROM d_labitems
  WHERE fluid = 'Blood'
    AND label IN ('White Blood Cells',
                  'Hemoglobin',
                  'Platelet Count',
                  'Creatinine',
                  'Urea Nitrogen',
                  'Sodium',
                  'Potassium',
                  'Glucose',
                  'Anion Gap',
                  'RDW')
),
labs_window AS (
  SELECT
    f.subject_id,
    f.hadm_id,
    li.label,
    le.valuenum
  FROM stg_analytic f
  INNER JOIN labevents le
    ON le.subject_id = f.subject_id
   AND le.hadm_id    = f.hadm_id
  INNER JOIN lab_items li ON li.itemid = le.itemid
  WHERE le.charttime >= f.icu_intime
    AND le.charttime <= f.icu_intime + INTERVAL '24 hours'
    AND le.valuenum IS NOT NULL
)
SELECT
  subject_id,
  hadm_id,
  MAX(CASE WHEN label = 'White Blood Cells'
           THEN valuenum END)          AS wbc_max,
  MIN(CASE WHEN label = 'Hemoglobin'
           THEN valuenum END)          AS hemoglobin_min,
  MIN(CASE WHEN label = 'Platelet Count'
           THEN valuenum END)          AS platelets_min,
  MAX(CASE WHEN label = 'Creatinine'
           THEN valuenum END)          AS creatinine_max,
  MAX(CASE WHEN label = 'Urea Nitrogen'
           THEN valuenum END)          AS bun_max,
  MIN(CASE WHEN label = 'Sodium'
           THEN valuenum END)          AS sodium_min,
  MAX(CASE WHEN label = 'Potassium'
           THEN valuenum END)          AS potassium_max,
  MAX(CASE WHEN label = 'Glucose'
           THEN valuenum END)          AS glucose_max,
  MAX(CASE WHEN label = 'Anion Gap'
           THEN valuenum END)          AS anion_gap_max,
  MAX(CASE WHEN label = 'RDW'
           THEN valuenum END)          AS rdw_max
FROM labs_window
GROUP BY subject_id, hadm_id;
")
```

## Coverage + distribution sanity check

```{r labs-coverage}
labs_coverage <- dbGetQuery(con, "
WITH cohort AS (
  SELECT COUNT(*) AS n_total FROM stg_analytic
)
SELECT
  (SELECT n_total FROM cohort)                               AS n_cohort,
  COUNT(*)                                                   AS n_with_any,
  ROUND(100.0 * COUNT(*)
        / (SELECT n_total FROM cohort), 1)                   AS pct_any,
  SUM(CASE WHEN wbc_max        IS NOT NULL THEN 1 END)       AS n_wbc,
  SUM(CASE WHEN hemoglobin_min IS NOT NULL THEN 1 END)       AS n_hb,
  SUM(CASE WHEN platelets_min  IS NOT NULL THEN 1 END)       AS n_plt,
  SUM(CASE WHEN creatinine_max IS NOT NULL THEN 1 END)       AS n_cr,
  SUM(CASE WHEN bun_max        IS NOT NULL THEN 1 END)       AS n_bun,
  SUM(CASE WHEN sodium_min     IS NOT NULL THEN 1 END)       AS n_na,
  SUM(CASE WHEN potassium_max  IS NOT NULL THEN 1 END)       AS n_k,
  SUM(CASE WHEN glucose_max    IS NOT NULL THEN 1 END)       AS n_glu,
  SUM(CASE WHEN anion_gap_max  IS NOT NULL THEN 1 END)       AS n_ag,
  SUM(CASE WHEN rdw_max        IS NOT NULL THEN 1 END)       AS n_rdw
FROM stg_labs_24h;
")
print(labs_coverage)

labs_summary <- dbGetQuery(con, "
SELECT
  ROUND(MEDIAN(wbc_max),        1) AS med_wbc,
  ROUND(MEDIAN(hemoglobin_min), 1) AS med_hb,
  ROUND(MEDIAN(platelets_min),  1) AS med_plt,
  ROUND(MEDIAN(creatinine_max), 2) AS med_cr,
  ROUND(MEDIAN(bun_max),        1) AS med_bun,
  ROUND(MEDIAN(sodium_min),     1) AS med_na,
  ROUND(MEDIAN(potassium_max),  2) AS med_k,
  ROUND(MEDIAN(glucose_max),    1) AS med_glu,
  ROUND(MEDIAN(anion_gap_max),  1) AS med_ag,
  ROUND(MEDIAN(rdw_max),        1) AS med_rdw
FROM stg_labs_24h;
")
print(labs_summary)
```

# First-24h physiology: vitals

Extract extremes of 6 vital signs within 24h of ICU intime from
`chartevents` (~30 GB). First query will take 5-10 minutes.

## Vital-sign itemid lookup

```{r explore-vital-items}
vital_items <- dbGetQuery(con, "
SELECT itemid, label, category, unitname
FROM d_items
WHERE itemid IN (
    220045,                    -- Heart Rate
    220179, 220050,            -- SBP (NIBP + arterial line)
    220180, 220051,            -- DBP (NIBP + arterial line)
    220181, 220052,            -- MBP (NIBP + arterial line)
    220210,                    -- Respiratory Rate
    220277,                    -- SpO2
    223761, 223762             -- Temperature F / C
  )
ORDER BY category, label;
")
print(vital_items)
```

## Extract 24h vital extremes

```{r extract-vitals-24h}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_vitals_24h AS
WITH vitals_window AS (
  SELECT
    f.subject_id,
    f.hadm_id,
    ce.itemid,
    ce.valuenum
  FROM stg_analytic f
  INNER JOIN chartevents ce
    ON ce.subject_id = f.subject_id
   AND ce.hadm_id    = f.hadm_id
  WHERE ce.itemid IN (
          220045,
          220179, 220050,
          220180, 220051,
          220181, 220052,
          220210,
          220277,
          223761, 223762
        )
    AND ce.charttime >= f.icu_intime
    AND ce.charttime <= f.icu_intime + INTERVAL '24 hours'
    AND ce.valuenum IS NOT NULL
)
SELECT
  subject_id,
  hadm_id,
  MAX(CASE WHEN itemid = 220045 THEN valuenum END)
                                          AS heart_rate_max,
  MIN(CASE WHEN itemid IN (220179, 220050)
           THEN valuenum END)             AS sbp_min,
  MIN(CASE WHEN itemid IN (220180, 220051)
           THEN valuenum END)             AS dbp_min,
  MIN(CASE WHEN itemid IN (220181, 220052)
           THEN valuenum END)             AS mbp_min,
  MAX(CASE WHEN itemid = 220210 THEN valuenum END)
                                          AS rr_max,
  MIN(CASE WHEN itemid = 220277 THEN valuenum END)
                                          AS spo2_min,
  MAX(CASE
        WHEN itemid = 223762 THEN valuenum
        WHEN itemid = 223761
          THEN (valuenum - 32) * 5.0 / 9.0
      END)                                AS temp_max_c
FROM vitals_window
GROUP BY subject_id, hadm_id;
")
```

## Vitals coverage + distribution

```{r vitals-coverage}
vitals_coverage <- dbGetQuery(con, "
WITH cohort AS (
  SELECT COUNT(*) AS n_total FROM stg_analytic
)
SELECT
  (SELECT n_total FROM cohort)                          AS n_cohort,
  COUNT(*)                                              AS n_with_any,
  ROUND(100.0 * COUNT(*)
        / (SELECT n_total FROM cohort), 1)              AS pct_any,
  SUM(CASE WHEN heart_rate_max IS NOT NULL THEN 1 END)  AS n_hr,
  SUM(CASE WHEN sbp_min        IS NOT NULL THEN 1 END)  AS n_sbp,
  SUM(CASE WHEN dbp_min        IS NOT NULL THEN 1 END)  AS n_dbp,
  SUM(CASE WHEN mbp_min        IS NOT NULL THEN 1 END)  AS n_mbp,
  SUM(CASE WHEN rr_max         IS NOT NULL THEN 1 END)  AS n_rr,
  SUM(CASE WHEN spo2_min       IS NOT NULL THEN 1 END)  AS n_spo2,
  SUM(CASE WHEN temp_max_c     IS NOT NULL THEN 1 END)  AS n_temp
FROM stg_vitals_24h;
")
print(vitals_coverage)

vitals_summary <- dbGetQuery(con, "
SELECT
  ROUND(MEDIAN(heart_rate_max), 1)  AS med_hr_max,
  ROUND(MEDIAN(sbp_min),        1)  AS med_sbp_min,
  ROUND(MEDIAN(dbp_min),        1)  AS med_dbp_min,
  ROUND(MEDIAN(mbp_min),        1)  AS med_mbp_min,
  ROUND(MEDIAN(rr_max),         1)  AS med_rr_max,
  ROUND(MEDIAN(spo2_min),       1)  AS med_spo2_min,
  ROUND(MEDIAN(temp_max_c),     2)  AS med_temp_max_c
FROM stg_vitals_24h;
")
print(vitals_summary)
```

# Build final merged analytic dataset

Merge baseline covariates (stg_analytic) + 24h labs + 24h vitals
+ survival outcomes into one patient-level table.

```{r build-analytic-final}
dbExecute(con, "
CREATE OR REPLACE TEMP TABLE stg_analytic_final AS
SELECT
  a.*,
  l.wbc_max, l.hemoglobin_min, l.platelets_min,
  l.creatinine_max, l.bun_max,
  l.sodium_min, l.potassium_max, l.glucose_max,
  l.anion_gap_max, l.rdw_max,
  v.heart_rate_max, v.sbp_min, v.dbp_min, v.mbp_min,
  v.rr_max, v.spo2_min, v.temp_max_c,
  s.dod, s.death_30d, s.time_to_event_30d, s.censored_30d
FROM stg_analytic a
LEFT JOIN stg_labs_24h   l USING (subject_id, hadm_id)
LEFT JOIN stg_vitals_24h v USING (subject_id, hadm_id)
LEFT JOIN stg_final_surv s USING (subject_id, hadm_id);
")

analytic_final <- dbGetQuery(con,
  "SELECT * FROM stg_analytic_final;")

cat("Rows:   ", nrow(analytic_final), "\n")
cat("Columns:", ncol(analytic_final), "\n\n")
cat("Column names:\n")
print(colnames(analytic_final))
```

## Per-variable missingness

```{r missingness}
miss_pct <- sapply(analytic_final, function(x)
  round(100 * sum(is.na(x)) / length(x), 1))
miss_tbl <- data.frame(
  variable = names(miss_pct),
  pct_missing = as.numeric(miss_pct)
)
miss_tbl <- miss_tbl[order(-miss_tbl$pct_missing), ]
print(miss_tbl)
```

## Export to CSV for downstream analysis

```{r export-csv}
out_path <- "analytic_dataset.csv"
write.csv(analytic_final, out_path, row.names = FALSE)

cat("Saved to:", normalizePath(out_path), "\n")
cat("File size:",
    round(file.info(out_path)$size / 1e6, 2),
    "MB\n")
```

```{r cleanup, eval=isTRUE(getOption("knitr.in.progress"))}
dbDisconnect(con, shutdown = TRUE)
```
